Complete Real-Time Liver Model Including Glisson's Capsule, Vascularization and Parenchyma

### 2013-05-28 Doc/Vrst2013/vrst2013.tex
FIX: caption bug and affiliations
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@574 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b
removing vascularization and global deformations
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@575 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\usepackage{amsmath}
%\usepackage{makeidx}
%\usepackage[pdftex]{graphicx} ... we have to use ps/eps
\usepackage{subfig}

@@

\numberofauthors{4}
\author{
% 1st. author
\alignauthor
Tom\'a\v{s} Golembiovsk\'y\\
       \affaddr{Faculty of Informatics}\\
       \affaddr{Masaryk University}\\
       \affaddr{Brno, Czech Republic}\\
       \email{nyoxi@ics.muni.cz}
% 2nd. author
\alignauthor
Igor Peterl\'ik\\
       \affaddr{Institut Hospitallo-Universitaire}\\
       \affaddr{Strasbourg, France}\\
       \email{...} % TODO
% 3rd. author
\alignauthor
Christian Duriez\\
       \affaddr{INRIA Lille}\\
       \affaddr{Lille, France}\\
       \email{...} % TODO
\and
% 4th. author
\alignauthor
Stephan\'e Cotin\\
       \affaddr{Institut Hospitallo-Universitaire}\\
       \affaddr{Strasbourg, France}\\
       \email{...} % TODO
}

@@

The mechanical importance of the vascular structures in liver is studied in~\cite{Peterlik2012}. It shows that the 
influence of vessels on the mechanical behaviour of the organ is significant, mainly if large deformations occur. 
Since a detailed modeling of the vessels would be extremely costly (mainly because of the small thickness of the vessel wall, 
the authors propose a composite model allowing for real-time simulation of entire liver with vascularization.

@@

In~\cite{Hollenstein2006}, a local influence of the capsule has been measured using a special aspiration device. The study was then repeated 
in vivo on human patients during the operation, confirming the mechanical importance of the membrane~\cite{Ahn2010,Nava2008}.
To our best knowledge, no attempt has been made to demonstrate the role of the Glisson's capsule on the global behaviour of 
the liver, mainly if the organ undergoes large deformations which is often the case during the surgery.

The main contribution of the paper is twofold: first, we present a real-time model of the liver, including
the parenchyma, vascularization and Glisson's capsule. The model is based on three different finite element representations for each constituent,
linked together via mechanical coupling. We show that the model mimics the \emph{local} experiments described in~\cite{Hollenstein2006}.

Second, we use the complete model of the liver to demonstrate the \emph{global} influence of the Glisson's
capsule via simulation: using a specific model of a porcine liver built from CT contrast-enhanced data, we show that there is a significant 
difference in the response of the model with and without the capsule in the case when the liver undergoes large deformations.

The paper is organized as follows: first we describe the proposed vascularized
liver model with capsule. Second, we validate our model in context of local deformations by reproducing the aspiration test 
described in~\cite{Hollenstein2006}. Finally, we demonstrate that in spite of its small thickness, the Glisson's 
capsule has a global influence on the liver undergoing large deformations.

%Then we show the correct behaviour of the
%model in context of \emph{local} deformations and finally we analyze the influence of
%the capsule on the \emph{global} deformations of the liver and show that despite the
%small thickness it plays an important role also in the \emph{global} context.

%Medical simulation is expected to play an integral part in many aspects of
%medical practice. From diagnosis, through planning or training to computer
%assisted intervention.
%Medical training application, like the ones for training laparoscopic
%surgery, have to operate in real-time and a good medical model has to be
%simple yet efficient to be really usable in this environment.
%
%Liver is largest internal organ formed mostly by a soft tissue (parenchyma)
%that is hyper-elastic and also visco-elastic. The parenchyma is interwoven
%by dense vascular network, this tree-like tubular structure is stiffer then
%the parenchyma. Finally the parenchyma is covered by very stiff collagenous
%membrane that preserves the integrity of the organ.
%
%Strangely, most of the work on liver modelling only considers the
%parenchyma and model the liver as single homogeneous object \TG{Refs}.
%Only recently
%Peterl\'{i}k et al. \cite{Peterlik2012} showed the importance of the vascular
%structures on the liver deformation.
%
%Another overlooked component is the Glisson's capsule.
%Biomechanical measurements report that the capsule has very small
%thickness (approx. 20 $\mu$m for porcine liver \cite{Umale2011}).
%On the other hand it has in three orders of magnitude higher stiffness than
%the parenchyma.
%Because of this it plays an important role in local
%deformations of the liver \cite{Ahn2010,Hollenstein2006}.
%It was reported that evaluation of the liver as a homogeneous material
%without capsule can lead to overestimation of local deformations up to the factor of 3
%\cite{Hollenstein2006}.
%
%While the influence on local deformation was reported, no experimental
%results show whether the capsule also plays an important role in global
%deformations of the liver. We propose a composite model that models the
%capsule with membrane FEM elements. We show that the model is able to
%simulate expected local behaviour reported in literature. Further more we
%show that the capsule plays also an important role in the deformation on
%global scale.

@@

In this section we describe the construction of a composite model
based on three components: tetrahedral FE model of the 
parenchyma, Tymoshenko-beam model for the vascular structures and finally 
triangular membrane elements used for the capsule.

@@

The parenchyma is modeled using corotational finite elements~\cite{Felippa2005}.
Although it relies on linear stress-strain relationship, large displacements including rotations are modelled correctly. 
While in the full non-linear formulation the stiffness matrix relates the forces $\Vec{f}$ and 
displacements $\Vec{u}$ as $\Vec{f} = \Mat{K}(\Vec{u})$, the corotational model 
requires the stiffness matrix $\Mat{K}_0$ of the system to be computed only once before the simulation begins. 
Then, in each step, the motion of each element $e$ is decomposed into rigid rotation $\Mat{R}^e$ and local deformation. 
The rotations are then used to update each local element stiffness matrix as $\Mat{R}\Mat{K}_0\Mat{R}^{\top}$
whereas the deformations are used to compute the linear strain in the local corotational frame.
%The stiffness matrix $\Mat{K}$ can be expressed in terms of the deformation
%$u$ and the force as $\Vec{f} = \Mat{K}(\Vec{u}) \Vec{u}$. 
%Where the
%components of $\Mat{K}$ depend on the orientation of each element in the
%current simulation step.
%the rigid component is  
% The corotational
%formulation allows us to deal with geometrical non-linearities but still
%relies on linear stress-strain relationship. Thus rotations and large
%deformations are modelled correctly. 
%In the corotational formulation the deformation
%of each element is expressed in the local frame of reference. 
There are several ways of computing the corotational frame for elements; we rely on
the geometrical method proposed in \cite{Nesme2005}.
% NOTE: This description of corotational method is very simplified and could be extended.

The model of the vascularization is based on linear beams 
with local frames of reference~\cite{Duriez2006}; in many aspects it's similar to the
corotational formulation described above. As such the model also handles geometric
non-linearities in the deformation. Through a specification of cross section and moments of inertia, 
the model can account for the specific properties of the blood vessels.

The beam-based model of vessels is mechanically coupled to the parenchyma as described in~\cite{Peterlik2012}. 
The coupling assumes that there is no relative motion between the vessels and the surrounding parenchyma. 
%The proposed model assumes no relative
%motion of vessels to the parenchyma. Segmented vascular network is
%discretized into the set of linear elements that are then projected into
%the tetrahedral mesh of the parenchyma. 
During the simulation, the nodes of linked beams are first displaced and rotated according to the actual motion of associated tetrahedra. 
As the deformation of beams results in mechanical response represented by forces and torques, these are propagated back to 
the tetrahedral FE model.

%a displacement of tetrahedral nodes are mapped  forces and torques resulting from the deformations of vessels by parenchyma are
%computed and propagated back onto the tetrahedral mesh.

@@

We use linear elastic material and employ the corotational formulation
for the CST elements.

@@

@@

\section{Experiments} %{{{
The model presented in the previous section was implemented in 
SOFA\footnote{www.sofa-framework.org} and a set of
numerical simulations was performed. First, local deformations were compared to the
results reported in literature to validate the method.
Second, to show that in spite of its very small thickness the membrane cannot be
neglected even in the context of global deformations and its overall
stiffness plays an important role, the model of
the complete liver was subjected to global deformations.

@@

\begin{figure*}
  \centering
    \includegraphics[width=.95\textwidth]{forceEvolution.pdf}
  \caption{\label{f:forceEvol}Evolution of forces during the simulation, where the supine data is deformed
towards the flank shape, and the stiffness of the capsule is increased from 0 to 8.22\,MPa after the initial equilibrium is achieved.}
\end{figure*}

\subsection{Global Deformations}
In the previous section we have shown that the composite model of parenchyma and 
capsule can be used to reproduce the \emph{local} behaviour of the tissue. In the following, 
we first present a real-time complete model of entire liver built from image data. 
Second, using this model we demonstrate the \emph{global} mechanical influence of the capsule when 
the model undergoes large deformations.

The complete model was constructed using contrast-enhanced CT data acquired on a female pig 
being positioned first in supine and then flank position. 
The liver and hepatic vein were segmented for both datasets  using semi-automatic methods available in ITKSnap~\footnote{www.itksnap.org}.
Further, an unstructured mesh was obtained from the segmented maps using CGAL~\footnote{www.cgal.org}. 
The semi-automatic method presented in~\cite{Peterlik2012} was 
then applied to the segmented map of the hepatic vein, resulting in linked-beam model. 
Finally, the model of the Glisson's capsule was constructed using the elements described in 
section~\ref{ss:capsuleModel} over the surface triangles extracted from the tetrahedral mesh. 
The complete model composed of 4266 tetrahedra, 188 beams and 1284 triangles was then 
used in simulation where external force representing the gravity was applied to the model, 
resulting in large deformations mainly in the area of lobes. The refresh rate of 22\,FPS 
was achieved on a desktop computer with Intel i7 CPU running at 3.4\,GHz and 8\,GB of memory. 
The simulation with gravity was also used as a first demonstration of the global influence of the capsule. 
In Fig.~\ref{f:global}a, the models with capsule (red) and without (yellow) are depicted after being 
deformed due to the gravity, showing a significant difference between the two models.

\begin{figure*}
  \centering
    \subfloat[]{\includegraphics[width=.32\textwidth]{gravity2.png}}
    \subfloat[]{\includegraphics[width=.32\textwidth]{registered.png}}
    \subfloat[]{\includegraphics[width=.32\textwidth]{distanceMap.png}}
    \caption{\label{f:global}%
      (a) Liver under gravity with (red) and without (yellow) capsule;
      (b) liver in supine (yellow) and flank (red) position;
      (c) Hausdorff distance map between registered version with and without capsule.}
\end{figure*}

In order to demonstrate the influence of the capsule on a real deformation of the liver, we 
performed a following protocol employing the supine and flank models built from the CT data. 
First, we performed a rigid registration of the two organs \wrt\ the root point of the hepatic vein. 
As shown in Fig.~\ref{f:global}b, the liver undergoes very large deformation when the body 
is turned from the supine to the flank position, exceeding 6\,cm for some parts of the organ. 
In order to induce similar deformation, we extracted feature pairs from the vascular system of the liver:
we selected points such as bifurcations which can be identified easily in both supine and flank 
data resulting in a set of correspondences. In the next step, we used these correspondences as a non-homogeneous
boundary conditions (imposed via penalty term) in order to perform a deformable registration of the supine
model onto the flank data. During the registration, the response forces acting in the points with prescribed displacements
were calculated. It is important to emphasize that the goal of the simulation was not 
to perform an accurate registration of the flank and supine data, but the simulation was used 
as an example of a real deformation of the organ.

The simulation was performed with the model composed of parenchyma,  vessels and capsule, however, zero stiffness was 
set for the last component.
Once the static equilibrium was achieved, the effect of the capsule was gradually switched 
on, \ie\ the stiffness of the membrane was slowly increased from 0 to 8.22\,MPa. The evolution of the force response in the 
points with prescribed displacements is depicted in Fig.~\ref{f:forceEvol} showing the change in the response as the effect
of the capsule is included. At the same time, the shape of the deformed liver with capsule was compared 
to the model without the capsule. A Hausdorff distance map for those cases is depicted in Fig.~\ref{f:global}c: in the red areas, the 
maximal difference exceeding 1.25\,cm was measured. We believe that this result confirms that the Glisson's capsule 
plays an important role in the global mechanical behaviour of the liver.

@@

\section{Conclusions} %{{{
In the paper, a complete model of the liver was presented taking into account 
three constituents of the organ, each being modelled with a different type 
of finite elements: tetrahedral corotational model used for the parenchyma, serially 
linked corotational beams based on Tymoshenko formulation employed to model the vascularization 
and finally CST elements used for the capsule. 
The model was first validated for local displacements: the aspiration test presented in~\cite{Hollenstein2006} 
was successfully reproduced, showing a good agreement between the simulation and experiments. 
Further, the model was used to demonstrate global influence of the capsule when the organ 
undergoes large deformations.

@@

%While the influence of the capsule on local deformations was previously
%studied in the literature, it's significance on the global scale
%deformations of the liver remained unexplored. We have performed a set of
%natural global deformations of the complete liver and shown that the
%capsule, despite its small thickness, plays a significant role also in
%global context.
%
%}}}

>>>>

\usepackage{caption}
\DeclareCaptionType{copyrightbox}
\usepackage{amsmath}
%\usepackage{makeidx}
%\usepackage[pdftex]{graphicx} ... we have to use ps/eps
\usepackage{subfig}

@@

\numberofauthors{4}
\author{
% 1st. author
\alignauthor
Tom\'a\v{s} Golembiovsk\'y\\
       \affaddr{Faculty of Informatics}\\
       \affaddr{Masaryk University}\\
       \affaddr{Brno, Czech Republic}\\
       \email{nyoxi@ics.muni.cz}
% 2nd. author
\alignauthor
Igor Peterl\'ik\\
       \affaddr{Institut Hospitalo-Universitaire}\\
       \affaddr{Strasbourg, France}\\
       \email{...} % TODO
% 3rd. author
\alignauthor
Christian Duriez\\
       \affaddr{INRIA}\\
       \affaddr{Lille, France}\\
       \email{...} % TODO
\and
% 4th. author
\alignauthor
Stephan\'e Cotin\\
       \affaddr{INRIA}\\
       \affaddr{Lille, France}\\
       \email{...} % TODO
}

@@

The mechanical importance of the vascular structures in liver is studied in~\cite{Peterlik2012}. It shows that the 
influence of vessels on the mechanical behaviour of the organ is significant. %, mainly if large deformations occur. 
%Since a detailed modeling of the vessels would be extremely costly (mainly because of the small thickness of the vessel wall, 
%the authors propose a composite model allowing for real-time simulation of entire liver with vascularization.
Since a detailed modeling of the vessels would be extremely costly (mainly because of the small thickness of the vessel wall, 
the authors propose a composite model allowing for real-time simulation.

@@

In~\cite{Hollenstein2006}, a local influence of the capsule has been measured using a special aspiration device. The study was then repeated 
in vivo on human patients during the operation, confirming the mechanical importance of the membrane~\cite{Ahn2010,Nava2008}.
%To our best knowledge, no attempt has been made to demonstrate the role of the Glisson's capsule on the global behaviour of 
%the liver, mainly if the organ undergoes large deformations which is often the case during the surgery.

In this paper we present preliminary results of our work on the complete liver
model. We present a real-time composite model accounting for parenchyma and
Glisson's capsule. The model is based on two different finite element
representations for each constituent coupled together.
%The main contribution of the paper is twofold: first, we present a real-time model of the liver, including
%the parenchyma, vascularization and Glisson's capsule. The model is based on three different finite element representations for each constituent,
%linked together via mechanical coupling.
We show that the model mimics the \emph{local} experiments described in~\cite{Hollenstein2006}.

%Second, we use the complete model of the liver to demonstrate the \emph{global} influence of the Glisson's
%capsule via simulation: using a specific model of a porcine liver built from CT contrast-enhanced data, we show that there is a significant 
%difference in the response of the model with and without the capsule in the case when the liver undergoes large deformations.

The paper is organized as follows: first we describe the proposed model of
liver with capsule. Second, we validate our model in context of local
deformations by reproducing the aspiration test described
in~\cite{Hollenstein2006}. We conclude by summarizing future steps towards the
complete liver model.
%Finally, we demonstrate that in spite of its small thickness, the Glisson's 
%capsule has a global influence on the liver undergoing large deformations.

@@

In this section we describe the construction of a composite model
based on two components: tetrahedral FE model of the 
parenchyma and
triangular membrane elements used for the capsule.

@@

The parenchyma is modeled using corotational finite elements~\cite{Felippa2005}.
Although it relies on linear stress-strain relationship, large displacements including rotations are modelled correctly. 
While in the full non-linear formulation the stiffness matrix relates the forces $\Vec{f}$ and 
displacements $\Vec{u}$ as $\Vec{f} = K(\Vec{u})$, the corotational model 
requires the stiffness matrix $\Mat{K}_0$ of the system to be computed only once before the simulation begins. 
Then, in each step, the motion of each element $e$ is decomposed into rigid rotation $\Mat{R}^e$ and local deformation. 
The rotations are then used to update each local element stiffness matrix as $\Mat{R}\Mat{K}_0\Mat{R}^{\top}$
whereas the deformations are used to compute the linear strain in the local corotational frame.
There are several ways of computing the corotational frame for elements; we rely on
the geometrical method proposed in \cite{Nesme2005}.
% NOTE: This description of corotational method is very simplified and could be extended.

%The model of the vascularization is based on linear beams 
%with local frames of reference~\cite{Duriez2006}; in many aspects it's similar to the
%corotational formulation described above. As such the model also handles geometric
%non-linearities in the deformation. Through a specification of cross section and moments of inertia, 
%the model can account for the specific properties of the blood vessels.

%The beam-based model of vessels is mechanically coupled to the parenchyma as described in~\cite{Peterlik2012}. 
%The coupling assumes that there is no relative motion between the vessels and the surrounding parenchyma. 
%During the simulation, the nodes of linked beams are first displaced and rotated according to the actual motion of associated tetrahedra. 
%As the deformation of beams results in mechanical response represented by forces and torques, these are propagated back to 
%the tetrahedral FE model.

@@

Similarly as with model of parenchyma we use linear elastic material and employ
the corotational formulation for the CST elements.

@@

\TG{We can also mention the mechanical mapping (like the one for vessels) as an
alternative approach. The result is same for conforming surface mesh.}

@@

\section{Experiments} %{{{
The model presented in the previous section was implemented in 
SOFA\footnote{www.sofa-framework.org} and a set of
numerical simulations was performed.
In this section we provide comparison of local deformations with the
results reported in literature to validate the method.
%Second, to show that in spite of its very small thickness the membrane cannot be
%neglected even in the context of global deformations and its overall
%stiffness plays an important role, the model of
%the complete liver was subjected to global deformations.

\TG{Here we have to decide wether to keep the values below as well as the Fig.
2 and specify the values used to match Hollenstein separately. Or if we take
just one set of values and redo the figure.}

@@

%\begin{figure*}
%  \centering
%    \includegraphics[width=.95\textwidth]{forceEvolution.pdf}
%  \caption{\label{f:forceEvol}Evolution of forces during the simulation, where the supine data is deformed
%towards the flank shape, and the stiffness of the capsule is increased from 0 to 8.22\,MPa after the initial equilibrium is achieved.}
%\end{figure*}

%\subsection{Global Deformations}
%In the previous section we have shown that the composite model of parenchyma and 
%capsule can be used to reproduce the \emph{local} behaviour of the tissue. In the following, 
%we first present a real-time complete model of entire liver built from image data. 
%Second, using this model we demonstrate the \emph{global} mechanical influence of the capsule when 
%the model undergoes large deformations. 
%
%The complete model was constructed using contrast-enhanced CT data acquired on a female pig 
%being positioned first in supine and then flank position. 
%The liver and hepatic vein were segmented for both datasets  using semi-automatic methods available in ITKSnap~\footnote{www.itksnap.org}.
%Further, an unstructured mesh was obtained from the segmented maps using CGAL~\footnote{www.cgal.org}. 
%The semi-automatic method presented in~\cite{Peterlik2012} was 
%then applied to the segmented map of the hepatic vein, resulting in linked-beam model. 
%Finally, the model of the Glisson's capsule was constructed using the elements described in 
%section~\ref{ss:capsuleModel} over the surface triangles extracted from the tetrahedral mesh. 
%The complete model composed of 4266 tetrahedra, 188 beams and 1284 triangles was then 
%used in simulation where external force representing the gravity was applied to the model, 
%resulting in large deformations mainly in the area of lobes. The refresh rate of 22\,FPS 
%was achieved on a desktop computer with Intel i7 CPU running at 3.4\,GHz and 8\,GB of memory. 
%The simulation with gravity was also used as a first demonstration of the global influence of the capsule. 
%In Fig.~\ref{f:global}a, the models with capsule (red) and without (yellow) are depicted after being 
%deformed due to the gravity, showing a significant difference between the two models.  
%
%\begin{figure*}
%  \centering
%    \subfloat[]{\includegraphics[width=.32\textwidth]{gravity2.png}}
%    \subfloat[]{\includegraphics[width=.32\textwidth]{registered.png}}
%    \subfloat[]{\includegraphics[width=.32\textwidth]{distanceMap.png}}
%    \caption{\label{f:global}%
%      (a) Liver under gravity with (red) and without (yellow) capsule;
%      (b) liver in supine (yellow) and flank (red) position;
%      (c) Hausdorff distance map between registered version with and without capsule.}
%\end{figure*}
%
%In order to demonstrate the influence of the capsule on a real deformation of the liver, we 
%performed a following protocol employing the supine and flank models built from the CT data. 
%First, we performed a rigid registration of the two organs \wrt\ the root point of the hepatic vein. 
%As shown in Fig.~\ref{f:global}b, the liver undergoes very large deformation when the body 
%is turned from the supine to the flank position, exceeding 6\,cm for some parts of the organ. 
%In order to induce similar deformation, we extracted feature pairs from the vascular system of the liver:
%we selected points such as bifurcations which can be identified easily in both supine and flank 
%data resulting in a set of correspondences. In the next step, we used these correspondences as a non-homogeneous
%boundary conditions (imposed via penalty term) in order to perform a deformable registration of the supine
%model onto the flank data. During the registration, the response forces acting in the points with prescribed displacements
%were calculated. It is important to emphasize that the goal of the simulation was not 
%to perform an accurate registration of the flank and supine data, but the simulation was used 
%as an example of a real deformation of the organ.
%
%The simulation was performed with the model composed of parenchyma,  vessels and capsule, however, zero stiffness was 
%set for the last component.
%Once the static equilibrium was achieved, the effect of the capsule was gradually switched 
%on, \ie\ the stiffness of the membrane was slowly increased from 0 to 8.22\,MPa. The evolution of the force response in the 
%points with prescribed displacements is depicted in Fig.~\ref{f:forceEvol} showing the change in the response as the effect
%of the capsule is included. At the same time, the shape of the deformed liver with capsule was compared 
%to the model without the capsule. A Hausdorff distance map for those cases is depicted in Fig.~\ref{f:global}c: in the red areas, the 
%maximal difference exceeding 1.25\,cm was measured. We believe that this result confirms that the Glisson's capsule 
%plays an important role in the global mechanical behaviour of the liver.

@@

\section{Future Work} %{{{

\TG{including vascularization; global deformation; stress the complexity of validation of global deformations}

\section{Conclusions} %{{{
\TG{TODO}
In the paper, a complete model of the liver was presented taking into account 
three constituents of the organ, each being modelled with a different type 
of finite elements: tetrahedral corotational model used for the parenchyma, serially 
linked corotational beams based on Tymoshenko formulation employed to model the vascularization 
and finally CST elements used for the capsule. 
The model was first validated for local displacements: the aspiration test presented in~\cite{Hollenstein2006} 
was successfully reproduced, showing a good agreement between the simulation and experiments. 
Further, the model was used to demonstrate global influence of the capsule when the organ 
undergoes large deformations.

@@
